---
title: "Exploring FAOSTAT package"
author: "Gabriel Antunes Daldegan"
date: "6/30/2017"
output: html_document
---

updated by TIMOTHY NGUYEN on 08/01/2017

```{r packages and constants}
library(readr)
library(dplyr)

bulk_path <- '/Users/timothy/Documents/FAOSTAT/data/bulk_data'
test_path <- '/Users/timothy/Documents/FAOSTAT/data/test_data'
output_path <- '/Users/timothy/Documents/FAOSTAT/data/output'
#dir_path <- '/home/nguyen/FAOSTAT/data'
```

# Objective
The goal is to retrieve data that satisfy the following query criteria:
- The entire Trade Matrix for 2007 to 2011. 
- All crops, all countries. 
- Import quantity (tonnes) is the variable of interest.

## Desired outcome
FAO has [a web client](http://www.fao.org/faostat/en/#data) that can be used to access their API. I have attempted to acquire our desired data through this client but ran into issues because the resulting dataset was too large for the client to handle. Thus, I had to manually download a comprehensive dataset that contains a detailed trade matrix via their 'bulk download' utility found in the right-hand side of [this page](http://www.fao.org/faostat/en/#data/TM).
```{r}
fao_data_path <- file.path(bulk_path, 'comp_trade_matrix.csv')
system.time(all_fao_data <- read_csv(fao_data_path)) 
head(all_fao_data)
desired_columns <- c(1:9, 52:61)
desired_data <- all_fao_data[all_fao_data$Element == 'Import Quantity', desired_columns]
head(desired_data)
```

FAO also offers a normalized version of this dataset:
```{r}
fao_data_norm_path <- file.path(bulk_path, 'comp_trade_matrix_norm.csv')
system.time(all_fao_data_norm <- read_csv(fao_data_norm_path))
head(all_fao_data_norm)
desired_years <- c(2007:2011)
desired_data_norm <-  all_fao_data_norm[all_fao_data_norm$Element == 'Import Quantity' & all_fao_data_norm$Year %in% desired_years,]
```

### Checking against web client
I have downloaded a .csv that contains data returned by query criteria: The entire Trade Matrix for 2007 to 2011. All crops, reporting country = afhanistan. Import quantity (tonnes) is the variable of interest.
```{r}
afg_path <- file.path(test_path, 'afg.csv')
true_afg <- read.csv(afg_path)

# discarding extraneous columns
standard_cols <- colnames(desired_data_norm)
true_afg <- true_afg %>%
  select(standard_cols)
# changing factor -> chars
true_afg <- data.frame(lapply(true_afg, as.character), stringsAsFactors=FALSE) 
```
Here, I subset the bulk_data according to the same criteria:
```{r}
test_afg <- desired_data_norm[desired_data_norm$`Reporter.Countries` == 'Afghanistan',]
test_afg <- data.frame(lapply(test_afg, as.character), stringsAsFactors=FALSE) # changing factor -> chars
```
comparing the two dfs reveals that they are not the same
```{r}
all.equal(test_afg,true_afg)
```
here are the nonequal rows
```{r}
row_equal <- function(row_a,row_b){
  return(sum(row_a == row_b) == 13)
}
non_equal_n <- c()
for (i in c(1:dim(test_afg)[1])){
  # print(row_equal(test_afg[i,], true_afg[i,]))
  if (!row_equal(test_afg[i,], true_afg[i,])) {
    non_equal_n <- c(non_equal_n, i)
  }
}
for (i in non_equal_n) {
  print('-----------------------')
  print(test_afg[i,])
  print(true_afg[i,])
}
```
it appears that the non equal rows are caused by character encoding errors; otherwise, the rows are identical. Thus, the downloaded df and the one I attained by subsetting the bulk data are identical.

Here are more tests:
data returned by query criteria: 
- The entire Trade Matrix for 2007 to 2011. 
- All crops, partner country == egypt. 
- Import quantity (tonnes) is the variable of interest.
```{r}
eg_path <- file.path(test_path, 'egypt.csv')
true_eg <- read.csv(eg_path)

# discarding extraneous columns
true_eg <- true_eg %>%
  select(standard_cols)
# changing factor -> chars
true_eg <- data.frame(lapply(true_eg, as.character), stringsAsFactors=FALSE) 
```
Here, I subset the bulk_data according to the same criteria:
```{r}
test_eg <- desired_data_norm[desired_data_norm$`Partner.Countries` == 'Egypt',]
test_eg <- data.frame(lapply(test_eg, as.character), stringsAsFactors=FALSE) # changing factor -> chars
```
comparing the two dfs reveals that they are not the same
```{r}
all.equal(test_eg,true_eg)
```
here are the nonequal rows
```{r}
eg_non_equal_n <- c()
for (i in c(1:dim(test_eg)[1])){
  if (!row_equal(test_eg[i,], true_eg[i,])) {
    eg_non_equal_n <- c(eg_non_equal_n, i)
  }
}
for (i in eg_non_equal_n) {
  print('-----------------------')
  print(test_eg[i,])
  print(true_eg[i,])
}
```
Again, it appears that the non equal rows are caused by character encoding errors; otherwise, the rows are identical. Thus, the downloaded df and the one I attained by subsetting the bulk data are identical. I think this is enough to say that subsetting the bulk data is a fruitful method.

### outputting desired data
```{r}
output_path <- file.path(output_path, 'trade_matrix.csv')
norm_output_path <- file.path(output_path, 'trade_matrix_normalized.csv')

write.csv(desired_data, output_path)
write.csv(desired_data_norm, norm_output_path)
```


# The package
Install and load FAOSTAT package

```{r}
#install.packages("FAOSTAT")
library(FAOSTAT)
```

Lets explore its documentation to better understand what it does

```{r}
vignette(topic = "FAOSTAT")
#demo(topic = "FAOSTATdemo")
#FAOsearch() # Function that allows to interactively choose variables of interest
```

Here I create two variables which will store the names and codes of all items (soybean, corn, cotton...)
```{r}
allItemNames <- FAOmetaTable$itemTable$itemName
allItemCodes <- FAOmetaTable$itemTable$itemCode
```

then we filter FAODataBase to build a data frame containing the variables of interest
```{r}
trade_matrix <- data.frame(
  varName = allItemNames,
  domainCode = "T",
  itemCode = allItemCodes,
  elementCode = 5610,
  stringsAsFactors = FALSE
)
```

and apply the data frame to the function that will download the data
```{r}
apple_df_lst <- with(
  trade_matrix[1:10,],
  getFAOtoSYB(
    name = varName,
    domainCode = domainCode,
    itemCode = itemCode,
    elementCode = elementCode,
    useCHMT = TRUE,
    outputFormat = "long"
  ))
```
